---
title: "Chapter 3: Inference for Simple Linear Regression"
subtitle: "Part 3: Section 3.6"
date: "Applied Statistics II"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "extra.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false
---
class: highlight-last-item

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width=9, fig.height=4,  
  out.width="90%",
  message = FALSE
)

```



```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(mosaic)
library(Stat2Data)
library(gridExtra)
library(plotly)
library(readr)
funneldata <- read_csv("http://people.kzoo.edu/enordmoe/math360/funneldata.csv", col_types = cols(trial = col_factor(levels = c("1","2"))))
funnel.hi <- filter(funneldata, height == 13)
funnel.hilo <- filter(funneldata, height == 13 | height == 10)
funnel.hilo <- mutate(funnel.hilo, height.hi = ifelse(height==13,1,0))
funneldata <- mutate(funneldata, height.hi = ifelse(height==13, 1, 0),
                   height.mid = ifelse(height == 11.5, 1, 0))
data("Perch")

```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("panelset"))
xaringanExtra::use_clipboard()
xaringanExtra::use_tile_view()
```


```{css echo=FALSE}
.highlight-last-item > ul > li,
.highlight-last-item > ol > li {
  opacity: 0.5;
}
.highlight-last-item > ul > li:last-of-type,
.highlight-last-item > ol > li:last-of-type {
  opacity: 1;
}
```

# Outline 

- Comparing Two Lines

--

- Nested $F$-test


---

### Funnel Data Model for Two Regression Lines

Recall the funnel data model for two regression lines:

$$
\text{tswirl} = \beta_0 + \beta_1 (\text{dist}) + \beta_2 (\text{funnel.hi}) + \beta_3 (\text{dist}\cdot \text{funnel.hi}) + \epsilon
$$

- Test for different slopes:

  - **t-test:** $H_0:\beta_3=0$ versus $H_a:\beta_3\neq 0$
  
--
  
- Test for different intercepts:

  - **t-test:** $H_0:\beta_2=0$ versus $H_a:\beta_2\neq 0$

--
  
- Test for different lines:

  - Test: $H_0: \beta_2=\beta_3=0$ versus $\text{At least one }\beta_i\neq 0$.
  
    - How?
    
---
### Second Order Model for SAT Data  Revisited

.panelset[

```{r, panelset = TRUE, fig.height = 3.5}
state_sat <- read.csv("http://people.kzoo.edu/enordmoe/math360/StateSAT.csv")
sat_second <- lm(SAT ~ Takers + Expend + I(Takers^2) 
                 + I(Expend^2) + I(Takers*Expend), data = state_sat)
summary(sat_second)
```
]

---
### Second Order Model for SAT Data  Revisited

**Inference Questions**

  - Do we need both quadratic terms?
  
--
  
  - Do we need the terms with Expend
  
--

**Q**: How to test just a subset of predictors?

--

**A**: Nested $F$-test


---

### Nested Models

**Definition**: If all of the predictors in Model A are also in a bigger Model B, we say that Model A is <br>**nested** in Model B.

--

**Example:**

  - $\text{tswirl} = \beta_0 + \beta_1 (\text{dist}) + \epsilon$ is *nested* in
  
  - $\text{tswirl} = \beta_0 + \beta_1 (\text{dist}) + \beta_2 (\text{funnel.hi}) + \beta_3 (\text{dist}\cdot \text{funnel.hi}) + \epsilon$

--

**Test for nested models:**

  - Do we really need the *extra* terms in Model B? 
  
  - Do the extra terms significantly improve on Model A?

---

### Nested $F$ test Procedure

1. Find how much "extra" variability is explained when the "new" terms being tested are added.

--

2. Divide by the number of new terms to get a mean square for the new part of the model.

--

3. Divide this mean square by the MSE for the "full" model to get an $F$ statistic.

--

4. Compare to an F distribution to find the $P$-value

---

## How much variability is explained by the "extra" predictors? 

- $\text{SSE}_\text{Full}$ - SS left unexplained by the full model

- $\text{SSE}_\text{Reduced}$ - SS left unexplained by the reduced model

  - Note  $\text{SSE}_\text{Full}\le \text{SSE}_\text{Reduced}$ for nested models.
  
--

<br>

**Difference in variability explained:**

$$\text{SSE}_\text{Reduced}-\text{SSE}_\text{Full}$$

$\Longrightarrow$ the difference is the "new" variability explained by the "extra" predictors in the full model.

---
## Nested $F$-test

Test:
- $H_0: \beta_i=0$ for a "subset" of predictors

- $H_a: \text{At least one }\beta_i\neq 0$ 

$$F=\frac{(\text{SSE}_\text{Reduced}-\text{SSE}_\text{Full})/(\#\text{ of extra predictors})}{\text{SSE}_\text{Full}/(n-k-1)}$$

--

or, equivalently


$$F=\frac{(\text{SSE}_\text{Reduced}-\text{SSE}_\text{Full})/(\#\text{ of extra predictors})}{\text{MSE}_\text{Full}}$$

- Compare to $F$ with $df_1=\#\text{ of extra predictors}$ and $df_2=\text{error df in full model}$.
---

### Example: Comparing Models with Nested F-Test in R



.panelset[
.panel[.panel-name[R code]
```{r nested1,eval = FALSE, fig.height=3}
sat_full <- lm(SAT ~ Takers + Expend + I(Takers^2) 
                 + I(Expend^2) + I(Takers*Expend), data = state_sat)
summary(sat_second)
anova(sat_second)
sat_reduced <- lm(SAT ~ Takers + (Takers^2), data = state_sat)
anova(sat_reduced)
anova(sat_reduced, sat_full)
```
]
.panel[.panel-name[full model output]
```{r nested2a, echo = FALSE}
sat_full <- lm(SAT ~ Takers + Expend + I(Takers^2) 
                 + I(Expend^2) + I(Takers*Expend), data = state_sat)
summary(sat_second)
```
]
.panel[.panel-name[full model anova]
```{r nested2b, echo = FALSE}
anova(sat_second)
```

]
.panel[.panel-name[reduced model output]
```{r nested3, echo = FALSE}
sat_reduced <- lm(SAT ~ Takers + (Takers^2), data = state_sat)
anova(sat_reduced)
```

]
.panel[.panel-name[nested F test]
```{r nested4, echo = FALSE, fig.height = 3}
anova(sat_reduced, sat_full)
```
]
]

---

### Example: Nested F Calculation

$$\begin{align}
F&=\frac{(\text{SSE}_\text{Reduced}-\text{SSE}_\text{Full})/(\#\text{ of extra predictors})}{\text{MSE}_\text{Full}}\\[2ex]
&=\frac{(64987-24669)/(48-44)}{24669/44}\\[2ex]
&=\frac{(40318)/4}{24669/44}= 17.978
\end{align}$$

```{r}
1-pf(17.978, 4,44)
```

---

### Special Cases for Nested F Test


- Test all predictors

  - "Usual" ANOVA $F$-test for the full model

- Test a single predictor (set one $\beta_i=0$)

  - "F test" equivalent of $t$- test
  
  ---
  
  ### Your Turn
  
  - Funnel Data
  
  - Car data
