---
title: "Chapter 3: Inference for Simple Linear Regression"
subtitle: "Part 3: Sections 2.4-2.5"
date: "Applied Statistics II"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "extra.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false
---
class: highlight-last-item

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width=9, fig.height=4,  
  out.width="90%",
  message = FALSE
)

```



```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(mosaic)
library(Stat2Data)
library(gridExtra)
library(plotly)
data("SpeciesArea")
data("CountyHealth")
data("LongJumpOlympics")
MidtermFinal <- read.csv("http://people.kzoo.edu/enordmoe/math360/MidtermFinal.csv")
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("panelset"))
xaringanExtra::use_clipboard()
xaringanExtra::use_tile_view()
```


```{css echo=FALSE}
.highlight-last-item > ul > li,
.highlight-last-item > ol > li {
  opacity: 0.5;
}
.highlight-last-item > ul > li:last-of-type,
.highlight-last-item > ol > li:last-of-type {
  opacity: 1;
}
```

# Outline 

- Intervals for Regression

--

- Confidence interval for $\mu_Y$

--

- Prediction interval for $Y$


  
---
## Accuracy of Predictions

### Example

Recall model for predicting Final grade from Midterm Grade.

$$\widehat{\mbox{Final}}=18.67+1.49\cdot\mbox{Midterm}$$


**Problem**: A student has a midterm grade of 41. What grade would we predict for the final?

$$\widehat{\mbox{Final}}=18.67+1.49 (41)=79.76$$

--

$\Longrightarrow$ How accurate is the prediction?

---

## Two Forms of Intervals for Regression 

1) Confidence interval for $\mu_Y$ (mean $Y$)

  - Where is the "true" line for a given $x$?
  
  - Where is the **average** $Y$ for the population of cases with a given $x$?
  
--

2) Prediction interval for individual $Y$

  - What are plausible $Y$ values for individuals with a given $x$ value?
  
---
## Confidence interval for $\mu_Y$ when $X=x^*$

$$\hat y \pm t^* \hat\sigma_{\epsilon} \sqrt{\frac 1n + \frac{(x^*-\bar x)^2}{\sum(x-\bar x)^2}}$$
<br>
The value of $t^*$ is the critical value for $t$ with $\mbox{df}=n-2$

--

**Question:** What factors affect the width of the interval?

---

## Prediction interval for individual $Y$ when $X=x^*$

$$\hat y \pm t^* \hat\sigma_{\epsilon} \sqrt{1+ \frac 1n + \frac{(x^*-\bar x)^2}{\sum(x-\bar x)^2}}$$

<br>

- This interval is much wider to account for individual variability in $Y$ values.

---

## Using R to calculate intervals for a new X


.panelset[

```{r, panelset = TRUE, fig.height = 3.5}
mod1 <- lm(Final ~ Midterm, data = MidtermFinal)
newx <- data.frame(Midterm = c(41))
predict(mod1, newdata = newx, interval = "confidence")
predict(mod1,  newdata = newx, interval = "prediction")
```
]

---

### Using R to calculate intervals: More Detail


- `se.fit` = $\mbox{SE}_{\hat Y}$ 

- `residual.scale` = $\hat\sigma_{\epsilon}$

.panelset[

```{r, panelset = TRUE, fig.height = 3.5}
mod1 <- lm(Final ~ Midterm, data = MidtermFinal)
newx <- data.frame(Midterm = c(42))
predict(mod1, newdata = newx, interval = "confidence", se.fit = TRUE)
```
]



---

### Visualizing the  Intervals


```{r, echo = FALSE, warning = FALSE}
mod1 <- lm(Final~Midterm,data=MidtermFinal)
MidFin_with_pred <- data.frame(MidtermFinal, predict(mod1, interval = 'prediction'))

gf_point(Final ~ Midterm, data=MidFin_with_pred) %>%
  gf_smooth(method = "lm") %>%
  gf_smooth(method = "lm", fill = ~"Confidence", alpha = 0.5, level = .95, se = TRUE) %>%
  gf_ribbon(lwr + upr ~ Midterm, fill = ~"Prediction",alpha = 0.2) %>%
  gf_refine(scale_fill_manual("Interval",values = c("green", "blue"))) %>%
  gf_theme(legend.position=c(.25, .85))

```

--

**Question:** Why isn't the prediction interval narrower near $\bar x$?

---

background-image: url(https://source.unsplash.com/-dChkgNLmp4)
background-position: top right
background-size: 300px

## Your turn
### Inference for the used car data

1. Fit the regression of Price against Mileage

2. Obtain an interval for the predicted average price of all cars with `Mileage` = 100 thousand.

3. Obtain an interval for the predicted price of an individual car for sale with `Mileage` = 100 thousand.

4. Which of your intervals is wider? Can you explain why?


---

## Summary 

- Two types of intervals associated with $\hat Y$.

- Prediction intervals (for $Y$) are always wider than confidence intervals (for $\mu_Y$).

- Value of $X=x^*$ determines both $\hat Y$ and the width of the intervals.

  - The more extreme $x^*$ is, the wider the intervals.

