---
title: "Chapter 3: Inference for Simple Linear Regression"
subtitle: "Part 2: Section 3.3"
date: "Applied Statistics II"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "extra.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false
---
class: highlight-last-item

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width=9, fig.height=4,  
  out.width="90%",
  message = FALSE
)

```



```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(mosaic)
library(Stat2Data)
library(gridExtra)
library(plotly)
library(readr)
funneldata <- read_csv("http://people.kzoo.edu/enordmoe/math360/funneldata.csv", col_types = cols(trial = col_factor(levels = c("1","2"))))
funnel.hi <- filter(funneldata, height == 13)
funnel.hilo <- filter(funneldata, height == 13 | height == 10)
funnel.hilo <- mutate(funnel.hilo, height.hi = ifelse(height==13,1,0))
funneldata <- mutate(funneldata, height.hi = ifelse(height==13, 1, 0),
                   height.mid = ifelse(height == 11.5, 1, 0))


```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("panelset"))
xaringanExtra::use_clipboard()
xaringanExtra::use_tile_view()
```


```{css echo=FALSE}
.highlight-last-item > ul > li,
.highlight-last-item > ol > li {
  opacity: 0.5;
}
.highlight-last-item > ul > li:last-of-type,
.highlight-last-item > ol > li:last-of-type {
  opacity: 1;
}
```

# Outline 

- "Dummy variable" predictors

  - As a single predictor

  - With a quantitative predictor
  
--
  
- Comparing two lines

  - Different intercepts
  
  - Different lines
  
--

- Extend to more than two lines

---

## Height as a Categorical Predictor

.pull-left[


- Use height $(x)$ to predict swirling time $(y)$

- Compare high (13 inches) vs low (10 inches)

]

--
.pull-right[

```{r, echo = FALSE}
gf_boxplot(tswirl ~ factor(height.hi), data = funnel.hilo) %>%
  gf_labs(title = "Swirling time by height", x = "0 = 10 inches, 1 = 13 inches" ) %>% 
gf_theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, color = "#166DA0", size = 32),
  )

```

]

---
### Two-Sample $t$ Test for Means (using pooled variances)

- Test for equal swirling times:

$$
\begin{gather}
H_0:\mu_1=\mu_2\\
H_a:\mu_1\neq\mu_2
\end{gather}
$$

- using pooled $t$-test:
$$t=\frac{\bar y_1-\bar y_2}{s_p\sqrt{\frac 1n_1 + \frac 1n_2}}$$

- where we assume constant variability $(\sigma_1=\sigma_2)$ and use the pooled standard deviation:
$$
s_p=\sqrt{\frac{(n_1-1)s_1^2+(n_2-1)s_2^2}{n_1+n_2-2}}\quad .
$$


---

### Two-sample $t$-test in R

.panelset[

```{r, panelset = TRUE, fig.height = 3.5}
t.test(tswirl ~ height.hi, var.equal = TRUE, data = funnel.hilo)
```
]

---

### Recast this as a dummy variable regression question

.pull-left[

- Regression model:

$$
\text{tswirl} = \beta_0 + \beta_1 X + \epsilon
$$
where
$$X=\begin{cases} 0 & \text{height } =10 \\1 & \text{height } =13 \end{cases}$$

- We call $X$ an **indicator** or **dummy** variable.

]
.pull-right[
```{r, echo = FALSE}
gf_point(tswirl ~ height.hi, data = funnel.hilo) +
  stat_summary(fun=mean, geom="line", color="#166DA0", lwd =2 )
```


]


---

### Dummy variable regression input and output

.panelset[

```{r, panelset = TRUE, fig.height = 3.5}
lm.dummy <- lm(tswirl ~ height.hi, data = funnel.hilo)
summary(lm.dummy)
```
]

---

### Comparing the t-test and regression output?

- Compare $t$ statistics and p-values.

- Interpret regression intercept.

- Interpret regression slope.

- Where does $s_p$ appear in the output?


---

## Quantitative plus indicator predictors
### Parallel lines model

.pull-left[

- Example:

  - $Y$ = swirling time
  
  - $X_1$ = distance up the channel
  
  - $X_2$ = height (low = 0, high = 1)
  
- Overall model

  - $Y=\beta_0+\beta_1 X_1+\beta_2 X_2 +\epsilon$
  
]

.pull-right[
  
- Identify models for each height

**low**: $X_2 = 0$ so 
$$\begin{align}Y=& \beta_0 + \beta_1 X_1+ \beta_2\cdot 0 + \epsilon \\[1ex]
=&\beta_1 X_1+  \epsilon\end{align}$$

  
**high**: $X_2 = 1$ so 
$$\begin{align}Y=&\beta_0 + \beta_1 X_1 + \beta_2\cdot 1 + \epsilon\\[1ex]
=& (\beta_0+\beta_2) + \beta_1 X_1 + \epsilon\end{align}$$
  
]

---

### Funnel Data Example: Parallel Lines Regression

.panelset[

```{r,panelset = TRUE, fig.height = 3.5}
lm.parallel = lm(tswirl ~ dist + height.hi, data = funnel.hilo)
summary(lm.parallel)
```

]

---

### Plotting the fitted parallel lines

.panelset[

```{r, panelset = TRUE, fig.height = 3.5}
gf_point(tswirl ~ dist, color = ~factor(height.hi), data = funnel.hilo) %>%
  gf_smooth(method = "lm")
```
]

---

### Comparing Parallel Regression Lines

- Is there a significant difference in the intercepts?

- The $t$-test for the `height.hi` (dummy variable) coefficient is a test of whether the intercepts differ:

$$H_0: \beta_2=0\\H_a: \beta_2\neq 0$$


---

### Assessing the Fit



.panelset[

```{r, panelset = TRUE, fig.height = 3.5}
# Optional: Code gets a bit messy to color residuals by height
# Don't worry about that
p1 <- gf_point(resid(lm.parallel)~fitted(lm.parallel), 
               color = ~factor(lm.parallel$model$height.hi)) %>%
  gf_labs(color = "Height")
p2 <- gf_qq(~resid(lm.parallel)) %>%
  gf_qqline()
grid.arrange(p1, p2, ncol = 2)
```

]


---

### Specify Fitted Models by Height

- Height = Low $(X_2=0)$:

$$
\widehat{\text{tswirl}} = 2.937 + 0.318 \cdot \text{dist}
$$

--

<br>

- Height = High $(X_2=1)$:

$$
\begin{align}
\widehat{\text{tswirl}} =& 2.937 + 0.318 \cdot \text{dist} + 11.211(1)\\[1ex]
=& 14.148 + 0.318 \cdot \text{dist} 
\end{align}
$$

---

### Interaction (Two Lines) Model

- Allow both intercepts and slopes to differ

- The two lines model:

$$
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1\cdot X_2 + \epsilon
$$

- $\beta_2$ models the difference in **intercepts**.

- $\beta_3$ models the difference in **slopes**.

---
### Interaction (Two Lines) Model for Funnel Data

- The complete model

$$
\text{tswirl} = \beta_0 + \beta_1 (\text{dist}) + \beta_2 (\text{funnel.hi}) + \beta_3 (\text{dist}\cdot \text{funnel.hi}) + \epsilon
$$

- Specifying the model for each height:

(1) Height = Low 
  
$$
\begin{align}
\text{tswirl} &= \beta_0 + \beta_1 (\text{dist}) + \beta_2 (0) + \beta_3 (\text{dist}\cdot 0) + \epsilon \\[1ex]
 &= \beta_0 + \beta_1 (\text{dist}) + \epsilon
\end{align}
$$
  
(2) Height = High 
  
$$
\begin{align}
\text{tswirl} &= \beta_0 + \beta_1 (\text{dist}) + \beta_2 (1) + \beta_3 (\text{dist}\cdot 1) + \epsilon \\[1ex]
 &= (\beta_0+\beta_2) + (\beta_1+\beta_3) (\text{dist}) + \epsilon
\end{align}
$$

---

### R Output for the Interaction Model


.panelset[

```{r, panelset = TRUE, fig.height = 3.5}
lm.2lines = lm(tswirl ~ dist + height.hi + dist * height.hi, data = funnel.hilo)
summary(lm.2lines)
```
]


---

## Using the R Output for the Interaction Model

### Use coefficient $t$-tests

- Test for different slopes:

  - $H_0: \beta_3=0\text{ versus } H_a: \beta_3\neq 0$
  
- Test for different intercepts:

  - $H_0: \beta_3=0\text{ versus } H_a: \beta_3\neq 0$
  

**Note:** Joint test for both requires a nested $F$-test (to come)  

---

### The Two Lines Models Can Be Generalized to More than Two Heights

.panelset[
.panel[.panel-name[3 parallel lines code]
```{r parmod_code,eval = FALSE, fig.height=3}
lm.parallel3 <- lm(tswirl ~ dist + factor(height), data = funneldata)
summary(lm.parallel3)
```
]
.panel[.panel-name[3 parallel lines output]
```{r parmod_code-out, ref.label="parmod_code", echo = FALSE}
```

]
.panel[.panel-name[3 lines code]
```{r lines_mod_code, eval = FALSE, fig.height = 3}
lm.3lines <- lm(tswirl ~ dist + factor(height) + dist:factor(height), 
             data = funneldata)
summary(lm.3lines)
```
]
.panel[.panel-name[3 lines output]
```{r lines_mod-out, ref.label = "lines_mod_code", echo=FALSE}
```
]
]

---
