---
title: "Chapter 5: One-Way ANOVA"
subtitle: "Section 5.6: Identifying Appropriate Transformations"
author: "MATH 360"
format:
  revealjs:
    chalkboard: true
    theme: [default, custom.scss]
    toc: false
    toc-depth: 1
    slideNumber: true
    html-math-method: mathjax
    incremental: false
    transition: fade
    preview-links: auto
    notes: true
title-slide-attributes:
  data-background-color: "#EA6820"
from: markdown+emoji
execute:
  echo: true
  cache: false
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(mosaic)
library(Stat2Data)
```


## Overview

- Transforming the response for ANOVA
  - Checking conditions
  - Diagnostic plot for spread to suggest transformation


##  {style="font-size: 60%;"}
### Example: Box Areas (1 of 3)

**Data:** `Boxes` dataset (constructed in Quarto file on Moodle)

- 20 boxes for each Width of 1, 4, 10, 12
- Random Length between 1 and 20 for each box
- Area = Length $\times$ Width

::: {.fragment}

```{r}
Boxes <- read.csv("http://people.kzoo.edu/enordmoe/math360/Boxes.csv")
favstats(areas ~ widths, data = Boxes) |> 
  select(widths, mean, sd, n)|>
  knitr::kable(digits = 2)
```

:::


::: {.fragment}

::: {.callout-warning}

**Big differences in SDs across groups** $\Longrightarrow$ equal variance condition is violated.

:::

:::


## Example: Box Areas (2 of 3)

```{r}
gf_boxplot(factor(widths) ~ areas, data = Boxes)
```

::: {.fragment}

:::{.callout-note}

* Boxplots show dramatically increasing spread as Width increases  
  — the SD for Width 20 is roughly 18 times larger than for Width 1.

:::

:::

---

## {.smaller}
### Diagnostic Plot for Addressing Unequal Variability


- Compute group means $(\bar{y}_i)$ and SDs $(s_i)$
- Plot $\log(s_i)$ versus $\log(\bar{y}_i)$
- Fit a line to the scatterplot and find the slope
- Try a transformation of $Y^p$, where $p = 1 - \text{slope}$

::: {.fragment}


| slope | $p = 1 - \text{slope}$ | Transformation | Response |
|------:|:----------------------:|:---------------|:--------:|
| $-1$ | $2$ | square | $Y^2$ |
| $0$ | $1$ | no change | $Y$ |
| $0.5$ | $0.5$ | square root | $\sqrt{Y}$ |
|$\Longrightarrow\quad$ $1$ | $0$ | logarithm | $\log(Y)$ |
| $2$ | $-1$ | reciprocal | $1/Y$ |

:::


## {.smaller}
### Diagnostic Plot for Boxes (Area by Width)

```{r}
# Compute group means and SDs, then plot log(s_i) vs log(y_bar_i)
stats <- favstats(areas ~ widths, data = Boxes)
gf_point(log(sd) ~ log(mean), data = stats) |>
  gf_lm()
```

::: {.fragment}


*The slope of the fitted line is approximately **1.01**.*

$$p = 1 - \text{slope} = 1 - 1.01 \approx 0$$

:::

::: {.fragment}

::: {.callout-tip}

**Conclusion:** $p \approx 0$ suggests a **log transformation**.

:::

:::

---

## Log(Area) versus Width

```{r}
favstats(log(areas) ~ widths, data = Boxes) |> 
  select(widths, mean, sd, n)|>
  knitr::kable(digits = 2)
```


::: {.fragment}

::: {.callout-note}

SDs across groups are now very similar $\Rightarrow$ the equal variance condition is satisfied after the log transformation.

:::

:::

## {.smaller}
### Example: Box Areas (3 of 3)

```{r}
gf_boxplot(factor(widths) ~ log(areas), data = Boxes)
```

::: {.fragment}

*After the log transformation, the four groups show comparable spread.  The boxplots have similar widths across all Width values.*

:::


##  {.smaller}
### Why Does the Log Transformation Work for Boxes?

Starting from the original relationship:

$$\text{Area} = \text{Width} \times \text{Length}$$

Taking logarithms of both sides:

$$\log(\text{Area}) = \log(\text{Width}) + \log(\text{Length})$$

| Term | Role in ANOVA |
|------|---------------|
| $\log(\text{Area})$ | Response variable |
| $\log(\text{Width})$ | Constant for each group (the "treatment effect") |
| $\log(\text{Length})$ | Random component (uniform length within each group) |



## {.smaller}
### Why Does the Log Transformation Work? (cont.)


Because $\text{Length}$ is drawn uniformly at random from the same distribution for every width group:

::: {.fragment}

- $\log(\text{Length})$ has the **same distribution** in every group
- $\log(\text{Width})$ is just a **constant shift** per group

:::

::: {.fragment}

* This is exactly the structure assumed by the one-way ANOVA model: equal variance across groups with group-specific means.

:::

::: {.fragment}

::: {.callout-note}

This diagnostic plot may also help suggest a transformation that addresses **skewness**, not just unequal variability.

:::

:::


## 
### Try it yourself

Apply the diagnostic plot to the **Olives** dataset (Problem 5.44) to investigate the dissipation of fenthion residue in olives over time.


<br>

::: {.fragment}


::: {.callout-warning}
**Caveat:** If the plot of $\log(s_i)$ versus $\log(\bar{y}_i)$ is **not very linear**, a power transformation might not be very helpful.
:::

:::


## Summary Choosing a Transformation

::: {.incremental}

- **Unequal spreads often occur because variability grows with the mean.**  
  When larger group means also have larger standard deviations, the equal variance assumption for ANOVA may fail.

- **A diagnostic plot reveals this relationship.**  
  Plot $\log(s_i)$ versus $\log(\bar{y}_i)$; the slope shows how the spread changes as the mean increases.

- **A power transformation can stabilize the spreads.**  
  Transform the response using $Y^p$ with $p = 1 - \text{slope}$ to make group standard deviations more nearly equal.

:::