---
title: "Two-Way ANOVA: From Additive Models to Interactions"
subtitle: "Math 360 Statistical Modeling — Class Notes"
format:
  html:
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: true
    theme: flatly
    embed-resources: true
    fig-width: 7
    fig-height: 4.5
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
library(mosaic)
library(tidyverse)
```

---


## The Gummy Bear Data

On Monday we launched gummy bears and recorded how far each one traveled under
combinations of two factors: **height** of the ramp (3 levels) and **position**
of the launcher on the ramp (3 levels), with 3 replications per treatment
combination: 27 launches in total.

### Read and inspect the data

> **Substitute your own data file path below.** Your data should have the same
> four columns as shown here: `height`, `position`, `replication`, and `distance`.

```{r read-data}
gummy <- read_csv("http://people.kzoo.edu/enordmoe/math360/gummybears2waydata.csv")
glimpse(gummy)
```

::: callout-important
**Before we go further:** `height` and `position` are currently stored as
integers (1, 2, 3). For ANOVA, they must be treated as **categorical** (factor)
variables, not numeric ones. We will convert them now.
:::

```{r factor-convert}
gummy <- gummy |>
  mutate(
    height   = factor(height,   labels = c("Low", "Medium", "High")),
    position = factor(position, labels = c("One", "Two",    "Three"))
  )
glimpse(gummy)
```

---

## The Two-Way ANOVA Setup

### Why two-way?

In one-way ANOVA we had a single categorical explanatory variable (factor) with
$k$ levels. Now we have **two factors**:
* $A$ (Height, with $I$ levels) 
* $B$ (Position, with $J$ levels).  

Each unique combination of a level of $A$ with a level of $B$ is called a treatment **cell**.  Here we have $I = 3$, $J = 3$, so $3 \times 3 = 9$ cells.

### Balanced design

Our experiment is **balanced**: each cell contains three replications $K=3$, for a grand total $n=IJK = (3)(3)(3)=27$. Balanced designs greatly simplify the algebra and interpretation.

### The two models we will build

| Model | Question it answers |
|---|---|
| **Additive (no interaction)** | Do *height* and *position* have separate, independent effects on distance? |
| **Interaction** | Does the effect of height *depend on*  the *position* of the  launcher (and vice versa)? |
 We will fit the simpler model first, check whether its assumptions are plausible, and then decide whether we need the richer interaction model.

---

## Exploratory Analysis

### Cell means and standard deviations

```{r favstats-cells}
favstats(distance ~ height + position, data = gummy)
```

It is helpful to arrange these as a grid. Here's a concise way to do this with Base R:

```{r}
tapply(gummy$distance, list(gummy$height, gummy$position), mean)
```


Scan across each row (fixed height, varying position) and down each column (fixed position, varying height).

- Do the row means increase or decrease consistently?
- Do the column means?
- Are any cells noticeably out of line with the rest?

### Design (main-effects) plot

The `plot.design()` function shows the **marginal means,** the mean distance averaged over all levels of the *other* factor, for each level of each factor.

```{r plot-design, fig.cap="Main-effects plot: Marginal Means for height and position"}
plot.design(distance ~ height + position, data = gummy)
```

**What to look for:** Large differences in marginal means suggest strong main effects. If the differences are much larger for one factor than the other, that factor likely has a stronger effect on distance.

### Interaction plot

The **interaction plot** reveals whether the two factors operate independently.

```{r interaction-plot, fig.cap="Interaction plot: mean distance by height, grouped by position"}
gf_summary(distance ~ height, color = ~position, 
           group = ~position, fun = mean, geom = "line", 
           xlab = "Height", ylab = "Mean Distance", 
           data = gummy)
```

**What to look for:**

- **Parallel lines** $\Rightarrow$ the factors act *additively*; the effect of height is the same regardless of position.
- **Non-parallel (crossing or converging) lines** $\Rightarrow$ evidence of an **interaction**; the effect of one factor changes depending on the level of the other.

::: callout-tip
**Discussion:** What do you see in the interaction plot for your group's gummy data?
Are the lines roughly parallel, or do they cross/converge?
:::

---

## The Two-Way Additive Model

### Model notation

The **two-way additive model** (also called the *main-effects model*) can be
written:

$$
Y_{ijk} = \mu + \alpha_i + \beta_j + \varepsilon_{ijk}
$$

where

| Symbol | Meaning |
|---|---|
| $Y_{ijk}$ | Observed distance for Height level $i$, Position level $j$, replication $k$ |
| $\mu$ | Overall (grand) mean |
| $\alpha_i$ | **Main effect of Height** level $i$ (deviation from $\mu$) |
| $\beta_j$ | **Main effect of Position** level $j$ (deviation from $\mu$) |
| $\varepsilon_{ijk}$ | Random error, assumed $\sim N(0, \sigma^2)$ |

The key assumption here is that the effect of Height ($\alpha_i$) is the
**same at every level of Position**, and vice versa. That is, the effect is "additive."

Once again, $\sum_i \alpha_i = 0$ and $\sum_j \beta_j = 0$, the effects add to 0 to allow estimation of the model.

### Fit the additive model

```{r mod-add}
mod_add <- aov(distance ~ height + position, data = gummy)
summary(mod_add)
```

### Reading the ANOVA table

The table partitions the total variability in distance into three pieces:

| Source | SS | df | MS | F |
|---|---|---|---|---|
| Height | $SS_A$ | $I-1 = 2$ | $MS_A$ | $MS_A / MS_E$ |
| Position | $SS_B$ | $J-1 = 2$ | $MS_B$ | $MS_B / MS_E$ |
| Residuals | $SS_E$ | $(n - I - J + 1)$ | $MS_E$ | — |

The *F*-statistic for each factor tests whether that factor's marginal means are all equal, **after accounting for the other factor**.

::: callout-note
**Degrees of freedom check:** With $n = 27$, $I = 3$, $I = 3$:

- Height: $df = 3 - 1 = 2$
- Position: $df = 3 - 1 = 2$
- Residuals: $df = 27 - 3 - 3 + 1 = 22$
- Total: $df = n-1 = 26$ ✓
:::

### Residual diagnostics for the additive model

Before accepting the model, we  check the three key conditions:

1. **Constant variance** (residuals spread equally across fitted values)
2. **Normality** of residuals
3. **Independence** (ideally, guaranteed by randomization of treatment order)

```{r diag-add, fig.cap="Residual diagnostics: additive model"}
#| layout-ncol: 2
gf_point(resid(mod_add) ~ fitted(mod_add),
         xlab = "Fitted Values", ylab = "Residuals",
         title = "Residuals vs. Fitted") |>
  gf_hline(yintercept = ~0, linetype = "dashed", color = "gray50")

gf_qq(~resid(mod_add), title = "Normal Q-Q Plot of Residuals") |>
  gf_qqline(color = "steelblue")
```

::: callout-tip
**Discussion:**  
Do the residuals fan out as fitted values increase (suggesting non-constant variance)?   
Does the Q-Q plot follow the reference line closely?
:::

### Model effects

The `model.tables()` function extracts the estimated effects $\hat\alpha_i$ and $\hat\beta_j$:

```{r effects-add}
model.tables(mod_add, type = "effects")
```

And the cell means predicted by the additive model:

```{r means-add}
model.tables(mod_add, type = "means")
```

---

## The Two-Way Interaction Model

### Why consider an interaction?

If the lines in the interaction plot are *not* parallel, the additive model is
**misspecified**; it will leave systematic patterns in the residuals and produce misleading tests. We need to add a term that captures how the two factors work together.

### Model notation

The **interaction model** adds a term $(\alpha\beta)_{ij}$ for each cell and thus requires replications in each cell:

$$
Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \varepsilon_{ijk}
$$

| Symbol | Meaning |
|---|---|
| $(\alpha\beta)_{ij}$ | Interaction effect for cell $(i,j)$: extra deviation beyond what the main effects alone predict |

This model has the constraint that $\sum_i (\alpha\beta)_{ij} = 0$ for all $j$, and
$\sum_j (\alpha\beta)_{ij} = 0$ for all $i$.

Conceptually: if there is **no interaction**, then $(\alpha\beta)_{ij} = 0$ for all cells and the model collapses back to the additive one which is "nested" in this model.

### Degrees of freedom

Adding the interaction term "uses up" additional degrees of freedom:

| Source | df |
|---|---|
| Height ($A$) | $I - 1 = 2$ |
| Position ($B$) | $J - 1 = 2$ |
| Height × Position | $(I-1)(J-1) = 4$ |
| Residuals | $IJ(K-1) = 9 \times 2 = 18$ |
| **Total** | $n-1=26$ |

Adding the interaction term uses up 4 more degrees of freedom, leaving fewer for estimating error, which makes it harder to detect real effects. So we only add it if the data give us good reason to.

### Fit the interaction model

As we saw earlier in the term, putting the `*` operator between two factors in the model specification automatically includes both main effects **and** their interaction:

```{r mod-int}
mod_int <- aov(distance ~ height * position, data = gummy)
summary(mod_int)
```

::: callout-note
Recall that `height * position` is shorthand for `height + position + height:position`. You can also write the interaction term explicitly as `height:position`.
:::

### Interpreting the interaction F-test

The `height:position` row tests $H_0$: all $(\alpha\beta)_{ij} = 0$
(i.e., the additive model is adequate) against $H_a$: at least one
interaction term is nonzero.

- **Large F, small p-value** ⇒ retain the interaction model.
- **Small F, large p-value** ⇒ the additive model may be sufficient.

::: callout-tip
**Discussion:** What does the interaction F-test tell you about your gummy bears data?
:::

### Residual diagnostics for the interaction model

```{r diag-int, fig.cap="Residual diagnostics: interaction model"}
#| layout-ncol: 2
gf_point(resid(mod_int) ~ fitted(mod_int),
         xlab = "Fitted Values", ylab = "Residuals",
         title = "Residuals vs. Fitted") |>
  gf_hline(yintercept = ~0, linetype = "dashed", color = "gray50")

gf_qq(~resid(mod_int), title = "Normal Q-Q Plot of Residuals") |>
  gf_qqline(color = "steelblue")
```

Compare these plots to those from the additive model. Does adding the interaction term clean up any systematic pattern in *your* residuals?

### Interaction effects

```{r effects-int}
model.tables(mod_int, type = "effects")
```

Scan the interaction effects table. Large positive values indicate cells where the observed mean *exceeds* what the main effects alone predict; large negative values indicate cells that *fall short*. Non-trivial values here are a sign that interaction is real and meaningful. But interaction effects may be hard to detect with just 3 replications per cell.

::: callout-tip
**Try it:** What happens if you replace `type = "effects"` with `type = "means"` in the `model.tables()` for the interaction model?
:::

---

## Model Comparison and Selection

To decide formally between the additive and interaction models, we can compare them directly. The interaction F-test in the `summary()` output already does this, but let's see it explicitly:

```{r model-compare}
anova(mod_add, mod_int)
```

The extra-sum-of-squares F-test asks: does adding the interaction term significantly reduce the residual sum of squares (taking into account the degrees of freedom required)?

### Which model should we use?

Use the following decision process:

1. **Look at the interaction plot.** Are the lines noticeably non-parallel?
2. **Check the interaction F-test p-value.** Is it below your chosen $\alpha$?
3. **Compare residual diagnostics.** Does the interaction model satisfy
   conditions better?

If the interaction is statistically significant *and* practically meaningful, use the interaction model. Otherwise, the simpler additive model is preferred.

---

## Post-Hoc Comparisons with Tukey's Method

When we reject the null hypothesis of equal means, we often want to know *which* groups differ. Tukey's HSD (Honestly Significant Difference) method controls the family-wise error rate across all pairwise comparisons.

::: callout-important
**When is Tukey's method appropriate here?**

- For the **additive model**: Tukey comparisons for `height` and `position`
  separately are straightforward and interpretable.
- For the **interaction model**: comparisons are most meaningful for the
  *cell means* (all Height × Position combinations). With 9 cells there are
  $\binom{9}{2} = 36$ pairwise comparisons. Use judgment about which to report since not all will be of interest.
:::

### Tukey comparisons for the additive model

```{r tukey-add}
TukeyHSD(mod_add)
```

```{r tukey-add-plot, fig.cap="Tukey 95% family-wise CI: additive model", fig.height=5}
par(las = 1, mar = c(5, 8, 4, 2)) # to allow space for labels
plot(TukeyHSD(mod_add))
par(las = 0, mar = c(5, 4, 4, 2))   # restore defaults
```

Intervals that **do not include zero** correspond to pairs of groups whose means differ significantly at the family-wise 5% level.

### Tukey comparisons for cell means (interaction model)

If the interaction is significant, it makes more sense to compare individual *cells* (Height × Position combinations):

```{r tukey-int-cells}
TukeyHSD(mod_int, which = "height:position")
```

```{r tukey-int-plot, fig.cap="Tukey 95% family-wise CI: interaction cells", fig.height=9}
par(las = 1, mar = c(5, 14, 4, 2))
plot(TukeyHSD(mod_int, which = "height:position"))
par(las = 0, mar = c(5, 4, 4, 2))
```

::: callout-tip
**Discussion:** Focus on the comparisons that are most interesting to your group. Which Height × Position combinations produce
significantly different distances?
:::

---

## Summary and Big Ideas

### The two-way ANOVA framework

| | Additive model | Interaction model |
|---|---|---|
| **Formula** | `aov(Y ~ A + B)` | `aov(Y ~ A * B)` |
| **Key assumption** | Effects of A and B are independent and simply add | The effect of A may differ across levels of B |
| **Interaction term** | Absent | $(\alpha\beta)_{ij}$ |
| **Residual df** | $n - I - J + 1$ | $IJ(K-1)$ |
| **Interaction plot** | Lines parallel | Lines non-parallel |
| **Post-hoc** | Tukey on A and B separately | Tukey on A:B cells |

### Key takeaways

1. **Two factors, two questions.** Two-way ANOVA simultaneously tests the main effects of two factors and, in the interaction model, whether those factors act independently.

2. **Start with the interaction model.** Fitting the interaction model first and testing whether the interaction term is needed is the safest strategy.

3. **Parallel lines = additive.** The interaction plot is the most intuitive
   diagnostic for deciding which model is appropriate.

4. **More parameters = harder to detect real effects** Adding an interaction costs $(I-1)(J-1)$ degrees of freedom. Only retain the interaction if the data support it.

5. **Tukey's method scales up.** The same family-wise error rate logic from one-way ANOVA applies here, just applied to a larger set of comparisons when the interaction is present.

6. **Conditions still matter.** The same three conditions required for one-way ANOVA (constant variance, normality, independence) apply to two-way ANOVA. Always check residual plots.

